---
layout: page
title: Introduction to dplyr
subtitle: clean and efficient data maninulation
minutes: 20
---
  
```{r knitr-options, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/')
```

> ## Learning Objectives
>
> * Introduce tbl_df data structure
> * Introduce the simplified syntax of dplyr
> * Introduce the concept of grouping in dplyr
> * Understand the concept of piping

## dplyr

[`dplyr`](https://github.com/hadley/dplyr) package offers simple, clear and 
efficient way of working with your data. The package makes the most common data
manipulation steps as fast and easy as possible by:

* Elucidating the most common data manipulation operations, so that your
  options are helpfully constrained when thinking about how to tackle a
  problem.
* Providing simple functions that correspond to the most common
  data manipulation verbs, so that you can easily translate your thoughts
  into code.
* Using efficient data storage backends, so that you spend as little time
  waiting for the computer as possible.

## From data frames to tbl_df 

```{r section-title-1, echo=FALSE, purl=TRUE}
### From data frames to tbl_df
```

dplyr introduces an extension to the regular data frame called `tbl` (a data
frame `tbl`). The main advantage to using a `tbl` over a regular data frame 
is the printing: `tbl` objects only print a few rows and all the columns that 
fit on one screen, describing the rest of it as text. For all practical 
purposes, `tbl` acts exactly like a regular data frame, i.e. syou can use the 
familiar `$` or `[` indexing notation.

Let's load the mammals survey data:

```{r load-data, purl=FALSE}
surveys <- read.csv('data/surveys.csv')
head(surveys)
```

Next, let's convert the data frame into a `tbl` object.

```{r, results='show', message=FALSE, purl=FALSE}
# If dplyr is not yet installed, uncomment the next line
# install.packages("dplyr")
library(dplyr)

surveys <- tbl_df(surveys)
# We don't need to use head() anymore, just printing surveys works
surveys
```

Note how the first row of print out (`Source: local data frame [35,549 x 9]`)
shows you also the dimensions of your table obect: 35,549 rows x 9 columns.

## dplyr syntax

```{r section-header-2, echo=FALSE, purl=TRUE}
### dplyr syntax
```

`dplyr` simplifies the syntax of many of the data manipulation operations that
you have been doing in R. For example, using the `surveys` data we can select 
all observations for females of the North American Deermouse 
([*Peromyscus maniculatus*](https://en.wikipedia.org/wiki/Peromyscus_maniculatus), 
see `"data/species.csv"` for the species information) made in January with:

```{r filter-example, results='show', purl=FALSE}
filter(surveys, month == 1, species_id == "PM" & sex == "F")
```

This is equivalent to the more conventional and verbose:

```{r old-style-filtering, results='show', purl=FALSE}
surveys[surveys$month == 1 & surveys$species_id == "PM" & surveys$sex == "F", ]
```

Another example of the simplified syntax is given how you sort your data frame
using `arrange()` function in `dplyr`:

```{r arrange-example, results='show', purl=FALSE}
arrange(surveys, year, month, day)
```

which is equivalent to:

```{r old-style-sorting, eval = FALSE, purl=FALSE}
surveys[order(surveys$year, surveys$month, surveys$day), ]
```

Selecting all columns except `year`, `month` and `day` is as simple as:

```{r select-example, results='show', purl=FALSE}
select(surveys, -year, -month, -day)
```

Renaming columns can be done usign `rename()`:

```{r rename-example, results='show', purl=FALSE}
rename(surveys, hf_len = hindfoot_length)
```

whereas renaming columns using base R has been somewhat painful:

```{r old-style-renaming, eval = FALSE, purl=FALSE}
colnames(surveys) <- gsub("hindfoot_length", "hf_len", colnames(surveys))
```

The following lesson topics will dive deeper into the different functionality
available in `dplyr`.

## Grouping

```{r section-header-3, echo=FALSE, purl=TRUE}
### Grouping
```

Another powerful feature of `dplyr` is the capability to combine functions
with the idea of "group by", repeating the operation individually on groups of 
observations within the dataset. In `dplyr`, you use the `group_by()` function 
to describe how to break a dataset down into groups of rows.

Let's calculate mean weight and hindfoot length for each species:

```{r group-by-example, results='show', purl=FALSE}
by_species <- group_by(surveys, species_id)
species_stats <- summarise(by_species,
  count = n(),
  weight = mean(weight, na.rm = TRUE),
  hindfoot_length = mean(hindfoot_length, na.rm = TRUE))
species_stats
```

Many species do not have recorded measures for `weight` and `hindfoot_length` at 
all (indicated by `NaN`s). Let's filter out these species

```{r, results='show', purl=TRUE}
filter(species_stats, !is.nan(weight) & !is.nan(hindfoot_length))
```


## Chaining (piping)

```{r section-header-4, echo=FALSE, purl=TRUE}
### Chaining (piping)
```

Most of the time your data manipulation will constiture of a sequence of 
operations applied on your data. You either have to do it step-by-step:

```{r save-to-vars, eval = FALSE, purl=FALSE}
a1 <- group_by(surveys, year, month, day)
a2 <- select(a1, record_id, weight, hindfoot_length)
a3 <- summarise(a2,
  mean_weight = mean(weight, na.rm = TRUE),
  mean_hf_length = mean(hindfoot_length, na.rm = TRUE))
a4 <- filter(a3, mean_weight > 30 | mean_hf_length > 30)
```

Or if you don't want to save the intermediate results, you need to wrap the 
function calls inside each other:

```{r nested-verbs, eval = FALSE, purl=FALSE}
filter(
  summarise(
    select(
      group_by(surveys, year, month, day),
      weight, hindfoot_length
    ),
    mean_weight = mean(weight, na.rm = TRUE),
    mean_hf_length = mean(hindfoot_length, na.rm = TRUE)
  ),
  mean_weight > 30 | mean_hf_length > 30
)
```

This is difficult to read because the order of the operations is from inside to 
out, and the arguments are a long way away from the function. To get around this 
problem, `dplyr` provides the `%>%` operator. `x %>% f(y)` turns into `f(x, y)` 
so you can use it to rewrite multiple operations so you can read from 
left-to-right, top-to-bottom:

```{r, piping, results='show', purl=FALSE}
surveys %>%
  group_by(year, month, day) %>%
  select(weight, hindfoot_length) %>%
  summarise(
    mean_weight = mean(weight, na.rm = TRUE),
    mean_hf_length = mean(hindfoot_length, na.rm = TRUE)
  ) %>%
  filter(mean_weight > 30 | mean_hf_length > 30)
```


