---
layout: topic
title: Introduction to dplyr
minutes: 45
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

> ## Learning Objectives
>
> * Familiarize participants with dplyr package
> * Introduce tbl_df data structure
> * Understand the data manipulation verbs used in dplyr
> * Understand the concept of piping

## dplyr

[`dplyr`](https://github.com/hadley/dplyr) package offers simple, clear and 
efficient way of working with your data. Repeating from the package
[vignette](https://github.com/hadley/dplyr/blob/master/vignettes/introduction.Rmd):

When working with data you must:

* Figure out what you want to do.
* Precisely describe what you want in the form of a computer program.
* Execute the code.

The `dplyr` package makes each of these steps as fast and easy as possible by:

* Elucidating the most common data manipulation operations, so that your
  options are helpfully constrained when thinking about how to tackle a
  problem.
* Providing simple functions that correspond to the most common
  data manipulation verbs, so that you can easily translate your thoughts
  into code.
* Using efficient data storage backends, so that you spend as little time
  waiting for the computer as possible.

## From data frames to tbl_df 

```{r, echo=FALSE, purl=TRUE}
### From data frames to tbl_df
```

dplyr introduces an extension to the regular data frame called `tbl` (a data
frame `tbl`). The main advantage to using a `tbl` over a regular data frame 
is the printing: `tbl` objects only print a few rows and all the columns that 
fit on one screen, describing the rest of it as text. Many functions in dplyr 
return `tbl` objects. For all practical purposes, `tbl` acts exactly like
a regular data frame, e.g. you can use the familiar `$` or `[` indexing 
notation. You can use `tbl_df()` to convert a regular data frame to `tbl` object 
and `as.data.frame()` to convert it back again.

Let's load the familiar mammals survey data:

```{r, purl=FALSE}
surveys <- read.csv('data/surveys.csv')
head(surveys)
```

Next, let's convert the data frame into a `tbl` object.

```{r, results='show', message=FALSE, purl=FALSE}
# Start by loading dplyr
library(dplyr)

surveys <- tbl_df(surveys)
# We don't need to use head() anymore, just printing surveys works
surveys
```

Note how the first row of print out (`Source: local data frame [35,549 x 9]`)
shows you also the dimensions of your table obect: 35,549 rows x 9 columns.

## Working with dplyr and the mammals data

```{r, echo=FALSE, purl=TRUE}
### Working with dplyr and the mammals data
```

Let's start exploring the capabilities of `dplyr` by looking at the basic single
table data manipulation verbs:

* `filter()` (and `slice()`)
* `arrange()`
* `select()` (and `rename()`)
* `distinct()`
* `mutate()` (and `transmute()`)
* `summarise()`
* `sample_n()` and `sample_frac()`

Let's look at each of these using the mammals data.

## Filter rows with `filter()`

```{r, echo=FALSE, purl=TRUE}
### Filter rows with `filter()`
```

`filter()` allows you to select a subset of the rows of a data frame. The first 
argument is the name of the data frame, and the second and subsequent are 
filtering expressions evaluated in the context of that data frame:

For example, using the `surveys` data we can select all observations for females 
of the North American Deermouse ([*Peromyscus maniculatus*](https://en.wikipedia.org/wiki/Peromyscus_maniculatus), see 
`"data/species.csv"` for the species information) made in January with:

```{r, results='show', purl=FALSE}
filter(surveys, month == 1, species_id == "PM", sex == "F")
```

This is equivalent to the more conventional and verbose:

```{r, eval = FALSE, purl=FALSE}
surveys[surveys$month == 1 & surveys$species_id == "PM", surveys$sex == "F", ]
```

`filter()` works similarly to `subset()` except that you can give it any number 
of filtering conditions which are joined together with `&` (not `&&` which is 
easy to do accidentally!).  You can use other boolean operators explicitly:

```{r, eval = FALSE, purl=FALSE}
filter(surveys, month == 1 | month == 2)
```

To select rows by position, use `slice()`:

```{r, results='show', purl=FALSE}
slice(surveys, 1:10)
```

## Arrange rows with `arrange()`

```{r, echo=FALSE, purl=TRUE}
### Arrange rows with `arrange()`
```

`arrange()` works similarly to `filter()` except that instead of filtering or 
selecting rows, it reorders them. It takes a data frame, and a set of column 
names (or more complicated expressions) to order by. If you provide more than 
one column name, each additional column will be used to break ties in the values 
of preceding columns:

```{r, results='show', purl=FALSE}
arrange(surveys, year, month, day)
```

Use `desc()` to order a column in descending order:

```{r, results='show', purl=FALSE}
arrange(surveys, desc(hindfoot_length))
```

`arrange()` is a straighforward wrapper around `order()` that requires less 
typing. The previous code is equivalent to:

```{r, eval = FALSE, purl=FALSE}
surveys[order(surveys$year, surveys$month, surveys$day), ]
surveys[order(desc(surveys$hindfoot_length)), ]
```

## Select columns with `select()`

```{r, echo=FALSE, purl=TRUE}
### Select columns with `select()`
```

Often you work with large datasets with many columns where only a few are 
actually of interest to you. `select()` allows you to rapidly zoom in on a 
useful subset using operations that usually only work on numeric variable 
positions:

```{r, results='show', purl=FALSE}
# Select columns by name
select(surveys, month, day, year)
# Select all columns between year and day (inclusive)
select(surveys, month:year)
# Select all columns except those from year to day (inclusive)
select(surveys, -(month:year))
```

This function works similarly to the `select` argument to the `base::subset()`. 
It's its own function in dplyr, because the dplyr philosophy is to have small 
functions that each do one thing well.

There are a number of helper functions you can use within `select()`, like 
`starts_with()`, `ends_with()`, `matches()` and `contains()`. These let you 
quickly match larger blocks of variable that meet some criterion. See `?select` 
for more details.

You can rename variables with `select()` by using named arguments:

```{r, results='show', purl=FALSE}
select(surveys, hf_len = hindfoot_length)
```

But because `select()` drops all the variables not explicitly mentioned, it's 
not that useful. Instead, use `rename()`:

```{r, eval=FALSE, results='show', purl=FALSE}
# Not evaluated
rename(surveys, hf_len = hindfoot_length)
```

## Extract distinct (unique) rows

```{r, echo=FALSE, purl=TRUE}
### Extract distinct (unique) rows
```

A common use of `select()` is to find out which values a set of variables takes. 
This is particularly useful in conjunction with the `distinct()` verb which only 
returns the unique values in a table.

```{r, results='show', purl=FALSE}
# Unique species
distinct(select(surveys, species_id))
# Which species are present in which plots
distinct(select(surveys, plot_id, species_id))
```

(This is very similar to `base::unique()` but should be much faster.)

## Add new columns with `mutate()`

```{r, echo=FALSE, purl=TRUE}
### Add new columns with `mutate()`
```

As well as selecting from the set of existing columns, it's often useful to add 
new columns that are functions of existing columns.  This is the job of 
`mutate()`:

```{r, results='show', purl=FALSE}
# Create new columns hindfoot_length_cm (hindfoot length in cm) and julian_time
# (day count from the first observation day 1977-07-16)
surveys_extended <- mutate(surveys,
              hindfoot_length_cm = hindfoot_length / 10,
              julian_time = julian(as.Date(paste0(year, "-", month, "-", day)),
                                   origin = as.Date("1977-07-16")))
# Show only selected columns
select(surveys_extended, record_id, year, month, day, species_id, 
       hindfoot_length_cm, julian_time)
```

If you only want to keep the new variables, use `transmute()`:

```{r, results='show', purl=FALSE}
surveys_extended <- transmute(surveys,
              hindfoot_length_cm = hindfoot_length / 10,
              julian_time = julian(as.Date(paste0(year, "-", month, "-", day)),
                                   origin = as.Date("1977-07-16")))
# Show the latest records
tail(surveys_extended, n=10)
```
