---
layout: topic
title: Starting with data
minutes: 20
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

> ## Learning Objectives
>
> * download the mammals data from Figshare (CSV files) and the data in memory
> using the survey table (`surveys.csv`) as an example
> * explore the structure and the content of the data in R
> * understand what are factors and how to manipulate them


# Presentation of the Survey Data

```{r, echo=FALSE, purl=TRUE}
# Presentation of the survey data
```

We are studying the species and weight of animals caught in plots in our study
area.  The dataset is stored as a `.csv` files in an online repository: each row 
holds information for a single animal, and the columns represent `record_id` 
`month`, `day`, `year`, `plot_id`, `species_id` (a 2 letter code, see the 
`species.csv` file for correspondance), `sex` ("M" for males and "F" for 
females), `hindfoot_length` (in millimeters), and `weight` (in grams).

The first few rows of the survey dataset look like this:

    "63","8","19","1977","3","DM","M","40"
    "64","8","19","1977","7","DM","M","48"
    "65","8","19","1977","4","DM","F","29"
    "66","8","19","1977","4","DM","F","46"
    "67","8","19","1977","7","DM","M","36"

To load our survey data, we need to locate the `surveys.csv` file. We will use
`read.csv()` to load into memory (as a `data.frame`) the content of the CSV
file.

* download [the data file`"surveys.csv"`](https://github.com/datacarpentry/datacarpentry/raw/master/data/biology/surveys.csv) from Figshsare
* put it in directory `"data"` within your working directory for these exercises
