---
layout: page
title: R dplyr ecology
subtitle: Data manipulation
minutes: 45
---


```{r setup, echo=FALSE, purl=FALSE}
source("setup.R")
```

> ## Learning Objectives
>
> * being able to select a column from a data frame
> * being able to filter rows on specific criteria
> * being able to order rows on specific criteria
> * being able to add a column as a function of existing columns

## Load data

```{r dplyr.init}
# if dplyr is not yet installed, uncomment the next line
# install.packages("dplyr")
library(dplyr)
```

Read in the data in comma separated file into a data frame.
```{r readSurveys}
surveys <- read.csv(file="data/surveys.csv")
```

## Selecting columns of data

In particular for larger datasets, it can be tricky to remember the column
number that corresponds to a particular variable. (is hindfoot length in column 7
or 9? oh, right... they are in column 8). In some cases, in which column the
variable will be can change if the script you are using adds or removes
columns. It is recommended to use column names to refer to a
particular variable, as it makes your code easier to read, and your intentions
clearer.

In the introduction to R, we have covered a first way of selecting columns, using the `$` sign.
If you want to look at the result, you can use head().

```{r dollar}
hindfoot_length_column <- surveys$hindfoot_length
# look at the data in the variable hindfoot_length_column
head(hindfoot_length_column)
```

Using dplyr, you can do operations on a particular column, by selecting it using the `select()` function.
The first argument is the data frame, the second argument the column name you want to select.
You can use `names(surveys)` or `colnames(surveys)` to remind yourself of the column names.

```{r select}
hindfoot_length_column <- select(surveys, hindfoot_length)
```

You can easily select multiple columns by adding the names as additional arguments.

```{r select_mult}
hindfoot_length_species_columns <- select(surveys, hindfoot_length, species_id)
```

## Filtering rows of data

When analyzing data, though, we often want to look at partial statistics, such
as the maximum value of a variable per species or the average value per plot.

One way to do this is to filter the data we want, and create a new temporary
data frame, using the `filter()` function. For instance, if we just want to look at
the animals of the species "DO":

```{r filter}
surveys_DO <- filter(surveys, species_id == "DO")
```

Filtering on multiple criteria (column names) can be done by adding extra filters:

```{r filter_mult}
surveys_DM <- filter(surveys, species_id == "DM", year >= "2001")
```



> ## Callout Box: Boolean operations and Filtering
>
> When you filter on a condition, you can select rows of the `data.frame`
> that match this condition. Let's see how this works with some examples using a
> vector:
>
> * First create a vector with some numbers:

```{r boolean1}
    u <- c(1, 4, 2, 5, 6, 3, 7)
```

> * We have already seen that we can use the positions of the elements to select
>   them (e.g., `u[3]` would return 2). We can also select elements by giving
>   them a vector of logicals (`TRUE` and `FALSE`):

```{r boolean-select, results='show'}
   u <- u[c(TRUE, FALSE, TRUE, TRUE, FALSE, FALSE, TRUE)]
```

> * When we ask R whether the content of a vector matches a given condition, it
>   returns a vector of the same size with each element being `TRUE` or `FALSE`,
>   depending if the elements of the vector match the condition:

```{r boolean2, results='show'}
    u < 3
```

> * If you want to select only the elements of the vector that match the
>   condition, we can combine these two concepts:

```{r boolean3}
    u[u < 3]
```

> * To negate the condition, we can use `!` in front of it:

```{r boolean5}
    u[! u < 3]
```

> * To know the number of elements that match the condition, we can use the
>   `sum()` function on the condition, and `mean()` will provide the proportion
>   of elements that match the condition

```{r boolean2b}
    sum(u < 3)
    mean(u < 3)
```

> * You can combine different conditions by using `&` (AND) and `|` (OR):

```{r boolean4}
    ## elements less than 3 OR greater or equal to 4:
    u[u < 3 | u >= 4]
    ## elements greater than 5 AND less than 1.
    ## As this is mathematically impossible, nothing matches this condition!
    u[u > 5 & u < 1 ]
    u[! (u > 5 & u < 1)]
    u[u > 5 & u < 7]
```

> ## Challenge: filtering rows on multiple factors

> To select rows from your data using `dplyr`, you can use these conditional
> statements with the `filter()` function. How many specimens of the species
> "DO" and "DM" were collected in 1977.

```{r filter_or, echo=FALSE, eval=FALSE}
    ## If you want to include multiple species, you can use a boolean operator | (OR):
    surveys_DO_DM <- filter(surveys, species_id == "DO" | species_id == "DM", year == "1977")
```

## Ordering rows according to a column

The function `arrange()` lets you reorder your data in ascending order:

```{r arrange}
surveys_order <- arrange(surveys, weight)
# have a look
head(surveys_order)
```

You can reverse the order using `desc()`, and order several columns at once (the
order in which you specify the columns matter):

```{r arrange_desc}
surveys_desc <- arrange(surveys, desc(year), desc(month), desc(day))
# have a look at the data
head(surveys_desc)
```

## Adding a column to our dataset

Sometimes, you may have to add a new column to your dataset that represents a
new variable, based on existing columns. You can do this using the function
`mutate()`.

Here we will add a column, converting the `hindfoot_length` from mm to cm:

```{r mutate}

surveys_plus <- mutate(surveys, hindfoot_length_cm = hindfoot_length / 10)
head(surveys_plus)

```

> ## Challenge adding column
>
> 1. Add a column calculating the ratio of hindfoot_lenght and weight.

```{r, echo=FALSE, eval=FALSE}
surveys_plus <- mutate(surveys, ratio = hindfoot_length / weight)
```

> 2. Because the weight and hindfoot_length is not known for all data, you can first filter the data to see some more meaningful results.

Select for rows where the weight is above 100 g and the hindfoot_length is longer than 10 mm.

```{r, echo=FALSE, eval=FALSE}
filter(surveys_plus, weight > 100, hindfoot_length > 10)
```
