---
layout: topic
title: Starting with data
minutes: 20
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```

> ## Learning Objectives
>
> * download the mammals data from Figshare (CSV files) and load the data in memory
> using the survey table (`surveys.csv`) as an example
> * explore the structure and the content of the data in R
> * understand what are factors and how to manipulate them

# Presentation of the Survey Data

```{r, echo=FALSE, purl=TRUE}
# Presentation of the survey data
```

We are studying the species and weight and hindfoot length of animals caught in plots in our study
area.  The dataset is stored as a `.csv` files in an (online repository)[http://figshare.com/articles/Portal_Project_Teaching_Database/1314459]: each row 
holds information for a single animal, and the columns represent `record_id` 
`month`, `day`, `year`, `plot_id`, `species_id` (a 2 letter code, see the 
`species.csv` file for correspondance), `sex` ("M" for males and "F" for 
females), `hindfoot_length` (in millimeters), and `weight` (in grams).

This is an example of what the survey dataset looks like:

    "63","8","19","1977","3","DM","M","35","40"
    "64","8","19","1977","7","DM","M","37","48"
    "65","8","19","1977","4","DM","F","34","29"
    "66","8","19","1977","4","DM","F","35","46"
    "67","8","19","1977","7","DM","M","35","36"

To load our survey data, we need to locate the `surveys.csv` file. We will use
`read.csv()` to load into memory (as a `data.frame`) the content of the CSV
file.

* download [the data file`"surveys.csv"`](http://files.figshare.com/1919744/surveys.csv) from Figshare
* put it in directory `"data"` within your working directory for these exercises

```{r download-data, echo=FALSE, purl=TRUE}
if (!file.exists("data/")) {
	dir.create("data")
}
if (!file.exists("data/surveys.csv")) {
    download.file("http://files.figshare.com/1919744/surveys.csv",
                  "data/surveys.csv")
}
if (!file.exists("data/species.csv")) {
    download.file("http://files.figshare.com/1919741/species.csv",
                  "data/species.csv")
}
```

Next, read the `surveys.csv` into memory using `read.csv()`:

```{r demo-data-load, eval=TRUE,  purl=FALSE}
surveys <- read.csv('data/surveys.csv')
```

__At this point, make sure all participants have the data loaded__

This statement doesn't produce any output because assignment doesn't display
anything. If we want to check that our data has been loaded, we can print the
variable's value: `surveys`

Wow... that was a lot of output. At least it means the data loaded
properly. Let's check the top (the first 6 lines) of this `data.frame` using the
function `head()`:

```{r, results='show', purl=FALSE}
head(surveys)
```
