---
layout: page
title: R dplyr ecology
subtitle: Data manipulation
minutes: 45
---

> ## Learning Objectives {.objectives}
>
> * being able to select a column from a data frame
> * being able to filter rows on specific criteria
> * being able to order rows on specific criteria
> * being able to add a column as a function of existing columns

## Load data

```{r check-data, echo=FALSE}
if (!file.exists("data/surveys.csv")) {
    download.file("http://files.figshare.com/1919744/surveys.csv",
                  "data/surveys.csv")
}
if (!file.exists("data/species.csv")) {
    download.file("http://files.figshare.com/1919741/species.csv",
                  "data/species.csv")
}
```

```{r dplyr.init}
# if dplyr is not yet installed, uncomment the next line
# install.packages("dplyr")
library(dplyr)
```

Read in the data in comma separated file into a data frame.
```{r readSurveys}
surveys <- read.csv(file="data/surveys.csv")
```

## Selecting columns of data

In particular for larger datasets, it can be tricky to remember the column
number that corresponds to a particular variable. (is hindfoot length in column 7
or 9? oh, right... they are in column 8). In some cases, in which column the
variable will be can change if the script you are using adds or removes
columns. It's therefore often better to use column names to refer to a
particular variable, and it makes your code easier to read and your intentions
clearer.

In the introduction to R, we have covered a first way of selecting columns, using the `$` sign.
If you want to look at the result, you can use head().

```{r dollar}
hindfoot_length_column <- surveys$hindfoot_length
# look at the data in the variable hindfoot_length_column
head(hindfoot_length_column)
```

Using dplyr, you can do operations on a particular column, by selecting it using the `select()` function. 
The first argument is the data frame, the second argument the column name you want to select.
You can use `names(surveys)` or `colnames(surveys)` to remind yourself of the column names.

```{r select}
hindfoot_length_column <- select(surveys, hindfoot_length)
```

You can easily select multiple columns by adding the names as additional arguments.

```{r select_mult}
hindfoot_length_species_columns <- select(surveys, hindfoot_length, species_id)
```

## Filtering rows of data

When analyzing data, though, we often want to look at partial statistics, such
as the maximum value of a variable per species or the average value per plot.

One way to do this is to filter the data we want, and create a new temporary
data frame, using the `filter()` function. For instance, if we just want to look at
the animals of the species "DO":

```{r filter, purl=FALSE}
surveys_DO <- filter(surveys, species_id == "DO")
```

Filtering on multiple criteria (column names) can be done by adding extra filters:

```{r filter_mult, purl=FALSE}
surveys_DM <- filter(surveys, species_id == "DM", year=="2002")
```

> ## Challenge filtering rows on multiple factors
> 
> 1. Filter rows on two different species e.g. both DO and DM
>
> Have a look at the [Introduction to dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html#filter-rows-with-filter) for a hint.

<!---
If you want to include e.g. multiple species, you can use a boolean operator OR ("|"):
```{r filter_or, purl=FALSE}
surveys_DO_DM <- filter(surveys, species_id == "DO" | species_id == "DM", year == "1977")
```
-->

## Ordering rows according to a column

In stead of filtering rows, you sometimes just want to reorder the data. 
This can be done using `arrange()`. It takes a data frame as first argument and a list of
column names. 

```{r arrange}
surveys_order <- arrange(surveys, year, month, day)
# have a look
head(surveys_order)
```

Default rows are ordered ascendingly, you can reverse this using `desc()`:

```{r arrange_desc}
surveys_desc <- arrange(surveys, desc(weight))
# have a look at the data
head(surveys_desc)
```

> ## Challenge ordering of rows
> 
> 1. Arrange the surveys so you see the most recent month first, ascendingly according to the day
> 

<!---
```{r arrange_challenge}
surveys_challenge <- arrange(surveys, desc(year), desc(month), day)
# have a look at the data
head(surveys_challenge)
```
-->

## Adding a column to our dataset

Sometimes, you may have to add a new column to your dataset that represents a
new variable, based on existing columns. You can do this using the function `mutate()`. 

Here we will add a column, converting the hindfoot_length from mm to cm:

```{r mutate}
surveys_plus <- mutate(surveys, hindfoot_length_cm = hindfoot_length / 10)
head(surveys_plus)

```

> ## Challenge adding column
> 
> 1. Add a column calculating the ratio of hindfoot_lenght and weight. 

<!---
>
> ```{r, results='show'}
> surveys_plus <- mutate(surveys, ratio = hindfoot_length / weight)
> ```
-->

> 2. Because the weight and hindfoot_length is not known for all data, you can first filter the data to see some more meaningful results.
Select for rows where the weight is above 100 g and the hindfoot_length is longer than 10 mm.

<!---
> ```{r}
> head(filter(surveys_plus, weight > 100, hindfoot_length > 10))
> ```
-->





