---
layout: page
title: Lesson Title
subtitle: Topic Title One
minutes: 10
---
> ## Learning Objectives {.objectives}
>
> * being able to subset for a level
> * being able to extract particular values from a `data.frame`
> * being able to add/remove columns and rows

# Data manipulation

## Load data

```{r check-data, echo=FALSE}
if (!file.exists("data/surveys.csv")) {
    download.file("http://files.figshare.com/1919744/surveys.csv",
                  "data/surveys.csv")
}
if (!file.exists("data/species.csv")) {
    download.file("http://files.figshare.com/1919741/species.csv",
                  "data/species.csv")
}
```

```{r dplyr.init}
# install.packages("dplyr")
library(dplyr)
```

```{r}
surveys <- read.csv(file="data/surveys.csv")
```

## Selecting columns of data

In particular for larger datasets, it can be tricky to remember the column
number that corresponds to a particular variable. (is hindfoot length in column 7
or 9? oh, right... they are in column 8). In some cases, in which column the
variable will be can change if the script you are using adds or removes
columns. It's therefore often better to use column names to refer to a
particular variable, and it makes your code easier to read and your intentions
clearer.

You can do operations on a particular column, by selecting it using the `select()` function. 
The first argument is the data frame, the second argument the column name you want to select.
You can use `names(surveys)` or `colnames(surveys)` to remind yourself of the column names.

```{r select}
select(surveys, hindfoot_length)
```

You can easily select multiple columns by adding the names as additional arguments.

```{r select_mult}
select(surveys, hindfoot_length, species_id)
```

## Filtering rows out of data

When analyzing data, though, we often want to look at partial statistics, such
as the maximum value of a variable per species or the average value per plot.

One way to do this is to filter the data we want, and create a new temporary
data frame, using the `filter()` function. For instance, if we just want to look at
the animals of the species "DO":

```{r filter, purl=FALSE}
surveys_DO <- filter(surveys, species_id == "DO")
```

Filtering on multiple criteria (column names) can be done by adding extra filters:

```{r filter_mult, purl=FALSE}
surveys_DM <- filter(surveys, species_id == "DM", year=="2002")
```





> ## Callout Box {.callout}
>
> An aside of some kind.

or an image from the `figs` directory:

![this is the image's title](fig/example.svg "this is the image's alt text")

> ## Challenge Title {.challenge}
>
> Description of a single challenge.
> There may be several challenges.



